Bandit Multiclass Linear Classification: Efficient Algorithms for the Separable Case

### 2019-01-22 ICML-submission/mistake-lower-bound-for-ignorant-algorithms.tex
Fixes

In this section, we prove a mistake lower bound for a family of algorithms
called \textit{ignorant algorithms}. Ignorant algorithms ignore the examples on
which they make mistakes. As an example, \textsc{Nearest-Neighbor Algorithm}
(Algorithm~\ref{algorithm:nearest-neighbor}) presented in
Appendix~\ref{section:nearest-neighbor-algorithm} is an ingnorant algorithm.
This assumption seems strong, but as we will explain below, it is actually
natural, and several recently proposed bandit classification algorithms that
achieve the $\sqrt{T}$-regret bound belong to this family, e.g.,
SOBA~\citep{Beygelzimer-Orabona-Zhang-2017},
OBAMA~\citep{Foster-Kale-Luo-Mohri-Sridharan-2018}. We will see that under the
$\gamma$-weak linearly separability assumption, if a bandit classification
algorithm wants to achieve a mistake bound better than
$\Theta\left(\min\{\sqrt{T}, (1/\gamma)^{\Theta(d)}\}\right)$, where $d$ is the
feature dimension, it must perform some meaningful update when it makes a
mistake.

The ignorant algorithm is formally defined as follows.
\begin{definition} \label{definition:ignorant-algorithm}
A bandit classification algorithm $\mathcal{A}$ is ignorant if for all
$t\in[T]$, $\widehat y_t\in \{1,2\}$ is drawn from some distribution that is
jointly determined by
\begin{enumerate}
\item $x_t$
\item $H_t\triangleq \left( (x_{\tau_1}, y_{\tau_1}), (x_{\tau_2}, y_{\tau_2}), \ldots, (x_{\tau_{n_t}}, y_{\tau_{n_t}}) \right)$,
where $1 \leq \tau_1 < \dots < \tau_{n_t} < t$, and $\{\tau_1, \ldots, \tau_{n_t}\}$
are the rounds in which the algorithm predicts correctly (that is,
$\{\tau_1, \ldots, \tau_{n_t}\} = \{1 \le \tau < t ~:~ \widehat y_\tau=y_\tau\}$).
\end{enumerate}
We use $p_t(\cdot|x_t)\in \{p\in \mathbb{R}_+^2: p^\top\mathbf{1}=1\}$ to denote
the distribution $\widehat y_t$ is drawn from. Note that we let the current
feature vector $x_t$ be an argument to $p_t$ to incorporate its dependency on
$x_t$. Therefore, for an ignorant algorithm, the function $p_t(\cdot|\cdot)$ is
only determined by $H_t$.
\end{definition}

An ignorant algorithm does not learn from mistakes in the following sense.
Suppose that an ignorant algorithm predicts incorrectly at time $t$ ($\widehat
y_t \neq y_t$). Then at time $t+1$, if it is given the same feature vector
$x_{t+1}=x_t$, it will predict the label with the same distribution
$p_{t+1}(\cdot|x_{t+1})=p_{t+1}(\cdot|x_t)=p_t(\cdot|x_t)$. Below we describe
the difficulty of learning from mistakes using a class of natural approaches
under the weak separability assumption. Consider an approach similar to OBAMA:
in each round $t$, we create a \textit{convex} loss function $\widehat \ell_t:
\mathbb{R}^{K\times d}\rightarrow \mathbb{R}$ that gives a loss value $\widehat
\ell_t(W)$ to all linear models $W\in\mathbb{R}^{K\times d}$ ($W=[w_1 \ldots
w_K]^\top$ for $w_1, \ldots, w_K$ defined in
Definition~\ref{definition:linear-separability}). The requirement for
$\widehat{\ell}_t$ to be convex is due to computational efficiency. To let
$\widehat{\ell}_t$ reflect the loss, if $\widehat y_t \neq y_t$, we want to
assign larger loss values $\widehat{\ell_t}(W)$ to the inconsistent models
$\mathcal{W}=\{W: \ip{w_{\widehat y_t}}{x_t} \geq \ip{w_i}{x_t}, \forall
i\in[K]\}$, while for other $W$'s, we want $\widehat{\ell_t}(W)$ to be smaller.
The following fact can be verified when $K\geq 3$, which suggests there is no
way to define the loss $\widehat \ell_t$ that satisfies the above properties
constraints: there exists $W_1, W_2, W_3$, where $W_2$ is in the line segment
between $W_1$ and $W_3$, such that $W_2\in \mathcal{W}$, while $W_1, W_3 \notin
\mathcal{W}$. On one hand, by convexity, we should have
$\widehat{\ell}_t(W_2)\leq \max\{\widehat{\ell}_t(W_1),
\widehat{\ell}_t(W_3)\}$; on the other hand, to penalize $\mathcal{W}$, we
should have $\widehat{\ell}_t(W_2) > \widehat{\ell}_t(W_1)$ and
$\widehat{\ell}_t(W_2) > \widehat{\ell}_t(W_3)$. Clearly, these two requirements
contradict each other.

@@

\begin{theorem}[Mistake Lower Bound for Ignorant Algorithms]
\label{theorem:ignorant_lower_bound}
Suppose $\mathcal{A}$ is an ignorant bandit classification algorithm. There
exists an adversary that can sequentially choose $(x_1, y_1), (x_2, y_2),
\ldots, (x_T, y_T) \in \mathbb{R}^d\times \{1,2\}$, such that $\forall t\in[T],
\|x_t\|\leq 1$, $\{(x_t, y_t)\}_{t=1}^T$ is $\gamma$-strongly linearly
separable, and the expected number of mistakes made by $\mathcal{A}$ is lower
bounded as
\begin{align*}
\Exp\left[ \sum_{t=1}^T \indicator{\widehat y_t \neq y_t} \right]
\geq \frac{1}{10}\min\left\{\sqrt{T}, \left(\frac{1}{160\gamma}\right)^{\frac{d-2}{4}} \right\} \; .
\end{align*}
\end{theorem}

Before proving the theorem, we need the following lemma.
\begin{lemma}
\label{lem:embed_d_gamma}
Let $\gamma \in (0,\frac 1 {160})$.
There exists vectors $v_1, \ldots, v_N, u_1, \ldots, u_N\in \mathbb{R}^d$, where $N = (\Omega(\frac{1}{\gamma}))^{(d-2)/2}$, such that:
$\| v_i \|_2 \leq 1$, $\| u_i \|_2 \leq 1$ for all $i$ in $[N]$, and for all $i, j$ in $[N]$,
\[ u_i \cdot v_j \begin{cases} \geq \gamma ,& i = j, \\ \leq -\gamma ,& i \neq j. \end{cases} \]
\end{lemma}
\begin{proof}
By Lemma 6 of~\cite{Long-1995}, there exists a set of $N =
(\frac{1}{2\sqrt{40\gamma}})^{d-2}$ $(d-1)$-dimensional vectors $z_1, \ldots,
z_N$ on the unit sphere $\mathbb{S}^{d-2}$, such that $\theta(z_i, z_j) \geq
\sqrt{40 \gamma}$. As $\cos\theta \leq 1-\theta^2/5$ for every $\theta \in
[0,\pi]$, this implies that
\[ z_i \cdot z_j \begin{cases} = 1 ,& i = j, \\ \leq 1- 8\gamma ,& i \neq j. \end{cases} \]

Define $v_i = (\frac12 z_i, \frac12)$, and $u_i = (\frac 12 z_i, -\frac
12(1-4\gamma))$ for all $i$ in $[N]$. It can be easily checked that for all $i$,
$\|v_i\| \leq 1$ and $\|u_i\| \leq 1$. In addition,
$$
u_i \cdot v_j
= \frac 1 4 z_i \cdot z_j - \frac {1-4\gamma} 4 \begin{cases} \geq \gamma ,& i = j, \\ \leq -\gamma ,& i \neq j. \end{cases}
$$
\end{proof}

\begin{proof}[Proof for Theorem~\ref{theorem:ignorant_lower_bound} (PartI: Mistake lower bound)]
We consider the strategy for the adversary described in Algorithm~\ref{algorithm:adversary-strategy}.
\begin{algorithm}[h]
\caption{\textsc{Adversary's strategy}}
\label{algorithm:adversary-strategy}
\begin{algorithmic}[1]
{
\STATE{\textbf{Define} $q_0=\max\left\{\frac{1}{\sqrt{T}}, \frac{1}{\sqrt{N}}\right\}$, where $N$ is defined in Lemma~\ref{lem:embed_d_gamma}. Also, let $v_1, \ldots, v_N$ be those defined in Lemma~\ref{lem:embed_d_gamma}.}
\STATE{\textbf{Initialize} $\textsc{phase}= 1$. }
\FOR{$t=1,\dots,\min\{T,N\}$}
    \IF{$\textsc{phase}=1$}
       \IF{$p_t(1|v_t)\geq 1-q_0$}
          \STATE{$(x_t, y_t)\leftarrow (v_t, 1)$}
       \ELSE
          \STATE{$(x_t, y_t)\leftarrow (v_t, 2)$}
          \STATE $\textsc{phase}\leftarrow 2$
       \ENDIF
    \ELSIF{$\textsc{phase}=2$}
       \STATE $(x_t, y_t)\leftarrow (x_{t-1}, y_{t-1})$.
    \ENDIF
\ENDFOR
\FOR{$t=\min\{T,N\}+1, \ldots, T$}
    \STATE $(x_t, y_t)\leftarrow (x_{t-1}, y_{t-1})$.
\ENDFOR

}
\end{algorithmic}
\end{algorithm}
For this strategy, we prove the mistake lower bound of any ignorant algorithm.
Define the indicators
\begin{align*}
A_t &= \indicator{\forall \tau\leq t, p_\tau(1|x_\tau)<1-q_0} \\
B_t &= \indicator{\exists \tau\leq t, p_\tau(1|x_\tau)\geq 1-q_0
 \text{\ and\ } \forall s\in[\tau,t), \widehat y_s \neq y_s}.
\end{align*}
Then we have
\begin{align}
& \mathbf{E}\left[\sum_{t=1}^{T} \indicator{\widehat y_t\neq y_t}\right] \nonumber \\
& \ge \mathbf{E}\left[\sum_{t=1}^{\min\{T,N\}} \indicator{\widehat y_t\neq y_t}\right] \nonumber \\
& \ge \mathbf{E}\left[\sum_{t=1}^{\min\{T,N\}} \indicator{\widehat y_t\neq y_t}A_t\right] \nonumber \\
& \qquad + \mathbf{E}\left[\sum_{t=1}^{\min\{T,N\}} \indicator{\widehat y_t\neq y_t}B_t\right] \nonumber \\
& = \mathbf{E}\left[\sum_{t=1}^{\min\{T,N\}} \mathbf{E}_t\left[\indicator{\widehat y_t\neq y_t}A_t\right]\right] + \mathbf{E}\left[\sum_{t=1}^{\min\{T,N\}} B_{t+1}\right] \nonumber \\
& \ge \mathbf{E}\left[\sum_{t=1}^{\min\{T,N\}} q_0 A_t\right] + \mathbf{E}\left[\sum_{t=1}^{\min\{T,N\}} B_{t+1}\right],
\label{eqn:mistake_lower_bound_temp}
\end{align}
where in the first inequality we use $A_t+B_t\leq 1$; in the equality, we use
$B_{t+1}=B_t\indicator{\widehat y_t\neq y_t}$ by $B_t$'s definition; in the last
inequality we use the fact that when $t\leq N$ and $A_t=1$, it must be $y_t=1$.
Now, denote $T_1 = \argmin_{\tau} \{ p_\tau(1|x_\tau) \ge 1-q_0 \}$ (if such
$\tau$ does not exist or is larger than $\min\{T,N\}$, we simply let
$T_1=\min\{T,N\}+1$). Then $A_t=1$ for all $t\leq T_1-1$.

Note that $B_{T_1}=1$, and for $t \ge T_1$ we have
$\Exp[B_{t+1}=0|B_t=1]=\Pr[\widehat y_t=y_t~|~B_t=1] = \Pr[\widehat y_t=2|B_t=1] \le
q_0$ because the algorithm is ignorant. Thus the second term in
\eqref{eqn:mistake_lower_bound_temp} can be calculated as

\begin{align*}
& \mathbf{E}\left[\sum_{t=1}^{\min\{T,N\}}B_{t+1}\right] \\
& \ge \mathbf{E}\left[\sum_{t=T_1}^{\min\{ T,N \}} (1-q_0)^{t-T_1+1} \right] \\
& = \frac{1-q_0}{q_0}\mathbf{E}\left[1-(1-q_0)^{\min\{T,N\}-T_1+1}\right] \; .
\end{align*}
Combining \eqref{eqn:mistake_lower_bound_temp} and the above inequalities, we get
\begin{multline}
\label{eqn:mistake_lower_bound_temp2}
\mathbf{E} \left[\sum_{t=1}^T \indicator{\widehat y_t\neq y_t}\right] \ge
\\ \mathbf{E}\left[q_0(T_1-1)+\frac{1-q_0}{q_0}\left(1-(1-q_0)^{\min\{T,N\}-T_1+1}\right)\right].
\end{multline}
In the case $T_1\geq \frac{1}{2}\min\{T,N\}+1$, the right-hand side of
\eqref{eqn:mistake_lower_bound_temp2} is lower bounded by
$\frac{1}{2}q_0\min\{T,N\}=\frac{1}{2}\min\{\sqrt{T}, \sqrt{N}\}$; in the case
$T_1< \frac{1}{2}\min\{T,N\}+1$, it is lower bounded by
\begin{align*}
& \frac{1-q_0}{q_0}\left(1-(1-q_0)^{\frac{1}{2}\min\{T,N\}}\right) \\
& = \frac{1-q_0}{q_0}\left(1-(1-q_0)^{\frac{1}{2q_0^2}}\right) \\
& \ge \frac{1-\frac{1}{\sqrt{2}}}{q_0}\left(1-\frac{1}{\sqrt{e}}\right) \\
& \ge \frac{1}{10}\min\{\sqrt{T}, \sqrt{N}\} \; .
\end{align*}
Combining both cases concludes the proof.
\end{proof}

\begin{proof}[Proof for Theorem~\ref{theorem:ignorant_lower_bound} (Part II: linear separability)]
Observe that in Algorithm~\ref{algorithm:adversary-strategy}, in the beginning
the adversary always set the label to $1$; once it sets $y_t=2$ at some $t$, it
lets $(x_s, y_s)=(x_t, y_t)=(v_t, 2)$ for all $s\geq t$. Consider $w_2=u_t$ and
$w_1=-u_t$, for $u_t$ as defined in Lemma~\ref{lem:embed_d_gamma}. Then by
Lemma~\ref{lem:embed_d_gamma}, we have $\ip{w_2}{x_s} = \ip{u_t}{v_t} \geq
\gamma, \forall s \geq t$ and $\ip{w_2}{x_s} = \ip{u_t}{v_s} \leq - \gamma,
\forall s< t$; also $\ip{w_1}{x_s} \leq -\gamma, \forall s \geq t$ and
$\ip{w_1}{x_s} \geq  \gamma, \forall s<t$. Thus, the example set is always
$\gamma$-strongly separable.
\end{proof}

>>>>

In this section, we prove a mistake lower bound for a family of algorithms
called \textit{ignorant algorithms}. Ignorant algorithms ignore the examples on
which they make mistakes. This assumption seems strong, but as we will explain
below, it is actually natural, and several recently proposed bandit
classification algorithms that achieve the $\sqrt{T}$-regret bound belong to
this family, e.g., SOBA~\citep{Beygelzimer-Orabona-Zhang-2017},
OBAMA~\citep{Foster-Kale-Luo-Mohri-Sridharan-2018}. Also,
\textsc{Nearest-Neighbor Algorithm} (Algorithm~\ref{algorithm:nearest-neighbor})
presented in Appendix~\ref{section:nearest-neighbor-algorithm} is an ingnorant
algorithm.

Under the assumption that the examples lie in in the unit ball of $\R^d$ and are
weakly linearly separable with margin $\gamma$, we show that any ignorant
algorithm must make $\Theta\left(\min\{\sqrt{T},
(1/\gamma)^{\Theta(d)}\}\right)$ mistakes in the worst case. In other words, an
algorithm that achieves a better mistake bound cannot ignore examples on which
it makes a mistake and it must make a meaningful update on such examples.

The ignorant algorithm is formally defined as follows.
\begin{definition} \label{definition:ignorant-algorithm}
A bandit classification algorithm $\mathcal{A}$ is ignorant if for all
$t\in[T]$, $\widehat y_t\in \{1,2\}$ is drawn from some distribution that is
jointly determined by
\begin{enumerate}
\item $x_t$
\item $H_t \triangleq \left( (x_{\tau_1}, y_{\tau_1}), (x_{\tau_2}, y_{\tau_2}), \ldots, (x_{\tau_{n_t}}, y_{\tau_{n_t}}) \right)$,
where $1 \le \tau_1 < \dots < \tau_{n_t} < t$, and $\{\tau_1, \ldots, \tau_{n_t}\}$
are the rounds in which the algorithm predicts correctly (that is,
$\{\tau_1, \ldots, \tau_{n_t}\} = \{1 \le \tau < t ~:~ \widehat y_\tau=y_\tau\}$).
\end{enumerate}
We use $p_t(\cdot|x_t)\in \{p\in \R_+^2 ~:~ p^\top\mathbf{1}=1\}$ to denote
the distribution $\widehat y_t$ is drawn from. Note that we let the current
feature vector $x_t$ be an argument to $p_t$ to incorporate its dependency on
$x_t$. Therefore, for an ignorant algorithm, the function $p_t(\cdot|\cdot)$ is
only determined by $H_t$.
\end{definition}

An ignorant algorithm does not learn from mistakes in the following sense.
Suppose that an ignorant algorithm predicts incorrectly at time $t$ ($\widehat
y_t \neq y_t$). Then at time $t+1$, if it is given the same feature vector
$x_{t+1}=x_t$, it will predict the label with the same distribution
$p_{t+1}(\cdot|x_{t+1})=p_{t+1}(\cdot|x_t)=p_t(\cdot|x_t)$. Below we describe
the difficulty of learning from mistakes using a class of natural approaches
under the weak separability assumption. Consider an approach similar to OBAMA:
in each round $t$, we create a \textit{convex} loss function $\widehat \ell_t:
\R^{K\times d} \to \R$ that gives a loss value $\widehat
\ell_t(W)$ to all linear models $W \in \R^{K\times d}$ ($W=[w_1 \ldots
w_K]^\top$ for $w_1, \ldots, w_K$ defined in
Definition~\ref{definition:linear-separability}). The requirement for
$\widehat{\ell}_t$ to be convex is due to computational efficiency. To let
$\widehat{\ell}_t$ reflect the loss, if $\widehat y_t \neq y_t$, we want to
assign larger loss values $\widehat{\ell_t}(W)$ to the inconsistent models
$\mathcal{W}=\{W: \ip{w_{\widehat y_t}}{x_t} \ge \ip{w_i}{x_t}, \forall
i\in[K]\}$, while for other $W$'s, we want $\widehat{\ell_t}(W)$ to be smaller.
The following fact can be verified when $K\ge 3$, which suggests there is no
way to define the loss $\widehat \ell_t$ that satisfies the above properties
constraints: there exists $W_1, W_2, W_3$, where $W_2$ is in the line segment
between $W_1$ and $W_3$, such that $W_2\in \mathcal{W}$, while $W_1, W_3 \notin
\mathcal{W}$. On one hand, by convexity, we should have
$\widehat{\ell}_t(W_2)\le \max\{\widehat{\ell}_t(W_1),
\widehat{\ell}_t(W_3)\}$; on the other hand, to penalize $\mathcal{W}$, we
should have $\widehat{\ell}_t(W_2) > \widehat{\ell}_t(W_1)$ and
$\widehat{\ell}_t(W_2) > \widehat{\ell}_t(W_3)$. Clearly, these two requirements
contradict each other.

@@

\begin{theorem}[Mistake Lower Bound for Ignorant Algorithms]
\label{theorem:ignorant_lower_bound}
Suppose $\mathcal{A}$ is an ignorant bandit classification algorithm. There
exists an adversary that can sequentially choose $(x_1, y_1), (x_2, y_2),
\ldots, (x_T, y_T) \in \R^d\times \{1,2\}$, such that $\forall t\in[T],
\|x_t\|\le 1$, $\{(x_t, y_t)\}_{t=1}^T$ is $\gamma$-strongly linearly
separable, and the expected number of mistakes made by $\mathcal{A}$ is lower
bounded as
\begin{align*}
\Exp\left[ \sum_{t=1}^T \indicator{\widehat y_t \neq y_t} \right]
\ge \frac{1}{10}\min\left\{\sqrt{T}, \left(\frac{1}{160\gamma}\right)^{\frac{d-2}{4}} \right\} \; .
\end{align*}
\end{theorem}

Before proving the theorem, we need the following lemma.

\begin{lemma}
\label{lemma:embed_d_gamma}
Let $\gamma \in (0,\frac 1 {160})$, let $d$ be a positive integer and let $N = (\Omega(\frac{1}{\gamma}))^{(d-2)/2}$.
There exist vectors $v_1, v_2, \dots, v_N, u_1, u_2, \dots, u_N \in \R^d$ such that for all $i, j \in \{1,2,\dots,N\}$
\begin{align*}
\norm{v_i} & \le 1 \, \\
\norm{u_i} & \le 1 \, \\
\ip{u_i}{v_j} & \ge \gamma & & \text{if $i=j$,} \\
\ip{u_i}{v_j} & \le -\gamma & & \text{if $i \neq j$.}
\end{align*}
\end{lemma}

\begin{proof}
By Lemma 6 of~\cite{Long-1995}, there exists a set of $N =
(\frac{1}{2\sqrt{40\gamma}})^{d-2}$ $(d-1)$-dimensional vectors $z_1, \ldots,
z_N$ on the unit sphere $\mathbb{S}^{d-2}$, such that $\theta(z_i, z_j) \ge
\sqrt{40 \gamma}$. As $\cos\theta \le 1-\theta^2/5$ for every $\theta \in
[0,\pi]$, this implies that
$$
\ip{z_i}{z_j} \begin{cases} = 1 ,& i = j, \\ \le 1- 8\gamma ,& i \neq j. \end{cases}
$$

Define $v_i = (\frac12 z_i, \frac12)$, and $u_i = (\frac{1}{2} z_i, -\frac{1}{2}(1-4\gamma))$ for all $i \in \{1,2,\dots,N\}$.
It can be easily checked that for all $i$,
$\norm{v_i} \le 1$ and $\norm{u_i} \le 1$. In addition,
$$
\ip{u_i}{v_j}
= \frac{1}{4} \ip{z_i}{z_j} - \frac {1-4\gamma} 4 \begin{cases} \ge \gamma ,& i = j, \\ \le -\gamma ,& i \neq j. \end{cases}
$$
\end{proof}

\begin{proof}[Proof for Theorem~\ref{theorem:ignorant_lower_bound} (PartI: Mistake lower bound)]
We consider the strategy for the adversary described in Algorithm~\ref{algorithm:adversary-strategy}.
\begin{algorithm}[h]
\caption{\textsc{Adversary's strategy}}
\label{algorithm:adversary-strategy}
\begin{algorithmic}[1]
{
\STATE{\textbf{Define} $q_0=\max\left\{\frac{1}{\sqrt{T}}, \frac{1}{\sqrt{N}}\right\}$, where $N$ is defined in Lemma~\ref{lemma:embed_d_gamma}. Also, let $v_1, \ldots, v_N$ be those defined in Lemma~\ref{lemma:embed_d_gamma}.}
\STATE{\textbf{Initialize} $\textsc{phase}= 1$. }
\FOR{$t=1,\dots,\min\{T,N\}$}
    \IF{$\textsc{phase}=1$}
       \IF{$p_t(1|v_t)\ge 1-q_0$}
          \STATE{$(x_t, y_t)\leftarrow (v_t, 1)$}
       \ELSE
          \STATE{$(x_t, y_t)\leftarrow (v_t, 2)$}
          \STATE $\textsc{phase}\leftarrow 2$
       \ENDIF
    \ELSIF{$\textsc{phase}=2$}
       \STATE $(x_t, y_t)\leftarrow (x_{t-1}, y_{t-1})$.
    \ENDIF
\ENDFOR
\FOR{$t=\min\{T,N\}+1, \ldots, T$}
    \STATE $(x_t, y_t)\leftarrow (x_{t-1}, y_{t-1})$.
\ENDFOR

}
\end{algorithmic}
\end{algorithm}
For this strategy, we prove the mistake lower bound of any ignorant algorithm.
Define the indicators
\begin{align*}
A_t &= \indicator{\forall \tau\le t, p_\tau(1|x_\tau)<1-q_0} \\
B_t &= \indicator{\exists \tau\le t, p_\tau(1|x_\tau)\ge 1-q_0
 \text{\ and\ } \forall s\in[\tau,t), \widehat y_s \neq y_s}.
\end{align*}
Then we have
\begin{align}
& \mathbf{E}\left[\sum_{t=1}^{T} \indicator{\widehat y_t\neq y_t}\right] \nonumber \\
& \ge \mathbf{E}\left[\sum_{t=1}^{\min\{T,N\}} \indicator{\widehat y_t\neq y_t}\right] \nonumber \\
& \ge \mathbf{E}\left[\sum_{t=1}^{\min\{T,N\}} \indicator{\widehat y_t\neq y_t}A_t\right] \nonumber \\
& \qquad + \mathbf{E}\left[\sum_{t=1}^{\min\{T,N\}} \indicator{\widehat y_t\neq y_t}B_t\right] \nonumber \\
& = \mathbf{E}\left[\sum_{t=1}^{\min\{T,N\}} \mathbf{E}_t\left[\indicator{\widehat y_t\neq y_t}A_t\right]\right] + \mathbf{E}\left[\sum_{t=1}^{\min\{T,N\}} B_{t+1}\right] \nonumber \\
& \ge \mathbf{E}\left[\sum_{t=1}^{\min\{T,N\}} q_0 A_t\right] + \mathbf{E}\left[\sum_{t=1}^{\min\{T,N\}} B_{t+1}\right],
\label{equation:mistake_lower_bound_temp}
\end{align}
where in the first inequality we use $A_t+B_t\le 1$; in the equality, we use
$B_{t+1}=B_t\indicator{\widehat y_t\neq y_t}$ by $B_t$'s definition; in the last
inequality we use the fact that when $t\le N$ and $A_t=1$, it must be $y_t=1$.
Now, denote $T_1 = \argmin_{\tau} \{ p_\tau(1|x_\tau) \ge 1-q_0 \}$ (if such
$\tau$ does not exist or is larger than $\min\{T,N\}$, we simply let
$T_1=\min\{T,N\}+1$). Then $A_t=1$ for all $t\le T_1-1$.

Note that $B_{T_1}=1$, and for $t \ge T_1$ we have
$\Exp[B_{t+1}=0|B_t=1]=\Pr[\widehat y_t=y_t~|~B_t=1] = \Pr[\widehat y_t=2|B_t=1] \le
q_0$ because the algorithm is ignorant. Thus the second term in
\eqref{equation:mistake_lower_bound_temp} can be calculated as

\begin{align*}
& \mathbf{E}\left[\sum_{t=1}^{\min\{T,N\}}B_{t+1}\right] \\
& \ge \mathbf{E}\left[\sum_{t=T_1}^{\min\{ T,N \}} (1-q_0)^{t-T_1+1} \right] \\
& = \frac{1-q_0}{q_0}\mathbf{E}\left[1-(1-q_0)^{\min\{T,N\}-T_1+1}\right] \; .
\end{align*}
Combining \eqref{equation:mistake_lower_bound_temp} and the above inequalities, we get
\begin{multline}
\label{equation:mistake_lower_bound_temp2}
\mathbf{E} \left[\sum_{t=1}^T \indicator{\widehat y_t\neq y_t}\right] \ge
\\ \mathbf{E}\left[q_0(T_1-1)+\frac{1-q_0}{q_0}\left(1-(1-q_0)^{\min\{T,N\}-T_1+1}\right)\right].
\end{multline}
In the case $T_1\ge \frac{1}{2}\min\{T,N\}+1$, the right-hand side of
\eqref{equation:mistake_lower_bound_temp2} is lower bounded by
$\frac{1}{2}q_0\min\{T,N\}=\frac{1}{2}\min\{\sqrt{T}, \sqrt{N}\}$; in the case
$T_1< \frac{1}{2}\min\{T,N\}+1$, it is lower bounded by
\begin{align*}
& \frac{1-q_0}{q_0}\left(1-(1-q_0)^{\frac{1}{2}\min\{T,N\}}\right) \\
& = \frac{1-q_0}{q_0}\left(1-(1-q_0)^{\frac{1}{2q_0^2}}\right) \\
& \ge \frac{1-\frac{1}{\sqrt{2}}}{q_0}\left(1-\frac{1}{\sqrt{e}}\right) \\
& \ge \frac{1}{10}\min\{\sqrt{T}, \sqrt{N}\} \; .
\end{align*}
Combining both cases concludes the proof.
\end{proof}

\begin{proof}[Proof for Theorem~\ref{theorem:ignorant_lower_bound} (Part II: linear separability)]
Observe that in Algorithm~\ref{algorithm:adversary-strategy}, in the beginning
the adversary always set the label to $1$; once it sets $y_t=2$ at some $t$, it
lets $(x_s, y_s)=(x_t, y_t)=(v_t, 2)$ for all $s\ge t$. Consider $w_2=u_t$ and
$w_1=-u_t$, for $u_t$ as defined in Lemma~\ref{lemma:embed_d_gamma}. Then by
Lemma~\ref{lemma:embed_d_gamma}, we have $\ip{w_2}{x_s} = \ip{u_t}{v_t} \ge
\gamma, \forall s \ge t$ and $\ip{w_2}{x_s} = \ip{u_t}{v_s} \le - \gamma,
\forall s< t$; also $\ip{w_1}{x_s} \le -\gamma, \forall s \ge t$ and
$\ip{w_1}{x_s} \ge  \gamma, \forall s<t$. Thus, the example set is always
$\gamma$-strongly separable.
\end{proof}
